Extended Early Adulthood and Attitude Change

### 2024-07-08 manuscript/manuscript.Rmd
drafting

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people have heightened probability of changing their opinions, beliefs, and political preferences. This heightened rate of change is often attributed to the social conditions of this life stage. In this paper, we explore whether substantial changes in the timing and ubiquity of major life-course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues. Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s. Instead, we find that other age groups have become more stable opinion holders."
---

@@

Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Other related explanations focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution. If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016]. If the mechanisms are principally cognitive and biological, rather than social, then these broader changes are unlikely to affect rates of social change in the same way.

In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005]. The "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.

Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of the year they were observed and their age at the time. Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant over time. They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents' reported opinions continued to change at roughly the same rate over time.

Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts. Life-course stages can potentially explain the divergence of opinion stability between the youngest age groups and older age groups over time, but they cannot explain the overall trajectory of increased stability.

@@

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. "Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These aggregate models of social imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains much more variance than period, the year of observation, for most items in the General Social Survey, suggesting a strong impressionable years pattern.

In recent years, researchers have attempted to use new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022]. This work has produced remarkably similar patterns with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019]. However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1950 are open to influence at the same rate as 20-year-olds in 2010 -- which has not been tested.

@@

Why does research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

## Biological Mechanisms

@@

It is also possible that there is something unique about human cognition during these periods making people ascribe more weight to the information acquired within them than to information acquired at other points. Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015]. Supporting this model is evidence that early formative periods have been observed not just in people, but across species. For example, in a study of social learning among birds, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.

@@

The thrust of the biological arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open changing dispositions at younger ages and relatively resistant at older ages. However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change. For that, we must turn attention to the structuring of the social world.

@@

Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct, characterized by sustained social learning and constant changes of social contexts. Here we briefly consider some of the ways that the "impressionable years" differ in ways that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.

First, people in this age group tend to move across social contexts at a much higher rates than people in other age groups. In the 11-year window from ages 14 to 25, in which people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information. People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons. Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups.

Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions. For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic]. Social networks also differ in this time period. Young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987]. It could be the case that these larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].

@@

While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on peoples' attitudes is contingent on their ability to incorporate it into their cultural repertoire. The social conditions of different life stages might make it easier or more difficult for people to act on these new sources of information in ways that facilitate opinion change.

Social networks not only shape peoples' exposure to new information, but they also affect the degree to which people adopt new beliefs and attitudes. People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks [@stark1980]. Consistent with this, mid-life is associated with denser, more homogenous networks [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004]. If young people exist in less dense, more heterogeneous social networks, it might be easier for them to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks. Similarly, if information flows less freely through weak-tie networks, these structures might allow respondents to limit their exposure to or close off parts of their lives to people who might not be supportive of new attitudes.

Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power. As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information [@eaton2009]. This potentially explains another phenomenon: that people appear to open up to changing their minds again as they enter later life and retirement, when these expectations presumably diminish [@danigelis2007].

@@

Given the distinct social conditions of the early-adult phase of life, there are reasons to believe that the "impressionable years" are principally an effect of how societies choose to structure the life course or the interaction between these life course stages and biological mechanisms. Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological or cognitive factors in explaining why we observe "impressionable years."

@@

First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts. Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce. In the United States, both the median age at first marriage and median age of first childbirth has consistently risen since the mid-20th Century [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts.

Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce. These transitions are also accompanied by heightened rates of geographic mobility, as well as movement between the workforce and education.

If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages. This, in turn, could have an effect on how long exploration, and therefore openness to durable change, lasts throughout the life-course. If this is the case, we should expect to see heightened levels of attitude change among age groups that, for earlier cohorts, displayed mainly stability. In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age. Specifically, we expect the attitude behavior of people aged 25-33 to more closely resemble middle-aged respondents in earlier cohorts [@krosnic1991; @alwin1991] but younger respondents in more recent cohorts. This produces our first hypothesis:

*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 25 to 33 than cohorts born in earlier years.*

@@

Rosenfeld's work suggests that the social conditions that shape where we spend our impressionable years might also shape how much exploration we can engage in. An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration. This assertion in turn engenders a second prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.

@@

*Hypothesis 2: More recent cohorts will show higher levels of opinion change in the 18 to 25 age range compared earlier cohorts.*

@@

While we have parimarily focused on changes in the timing and structure of life course stages over the past 70 years, that time frame has also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure. We briefly consider three broader changes here.

First, the number of people pursuing and completing higher education increased significantly over the window studied here [@schofer2005; @hout2012]. Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007]. The specific mechanism that links educational attainment and attitude stability is unclear, with some perspectives suggesting educational attainment improves cognitive function and memory while other perspectives suggest that educational attainment makes people pay more attention to current events. It could also be that education cultivates in people an expectation that they hold opinions in the first place. Regardless of the specific mechanism, a population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.

Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding. Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes opinion-holding easier.

Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008]. People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020]. While clarity aroud political issue positions is most commonly studied, the same forces can also align other social beliefs and identities. This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992]. Geographic and social sorting into lifestyle "enclaves" also reinforce these processes, continually providing clearer signals about which opinions people should hold [@dellaposta2015].

Broadly speaking, these three forces suggest that opinions should broadly become more stable over time, regardless of life-course stages. Additionally, opinions on certain issues might become more solidified the longer issues remain in the public, as social movements and organizations become solified around them.

@@

Political and Social Attitudes: Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts. Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, with respondents coming from the same general population. To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years. Both the ANES and GSS are designed to generalize to the non-institutionalized U.S. adult (18+) population, and therefore can be thought of as tapping the same general population. We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20. We evaluated an additional panel conducted by the ANES in 2008-10, but no questions overlapped this panel and any other panel.

>>>>

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences. This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect biological or cognitive differences between young adults and older adults. In this paper, we explore whether substantial changes in the timing and ubiquity of major life-course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues. Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s. Instead, we find that other age groups have become more stable opinion holders. These aggregate trends obscure two countervailing patterns: decreases in stability due to life-course changes among younger adults and increases in stability across all age groups due to increases in educational attainment, as well as a broader stability trend that remains unexplained. Our results suggest a braod role in social contexts in shaping attitude stability."
---

@@

Why people in these life stages appear to show more frequent opinion change relative to older people, however, is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Related explanations conversely focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

It is likely true that both social and cognitive/biological mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution. If the mechanisms are principally cognitive and biological, rather than social, then broader changes in the structure of different life stages are unlikely to affect rates of social change in the same way. If, conversely, openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016].

Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005]. Rates of high school and college completion have also increased significantly. As a result, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.

In this paper, we test whether these broader shifts in social structure and the timing of life course transitions have affected the rate at which young people change their attitudes about political and social issues. Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we estimate the amount of change in opinions as a function of age and cohort. Consistent with the "impressionable years" hypothesis, we find that people in age ranges associated with this "early adulthood" life stage (18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant over time. They emerged over the second half of 20th Century as more recent cohorts exhibited more stable opinions in middle and old age compared to their predecessors, while early adult respondents' reported opinions continued to change at roughly the same rate as in earlier cohorts.

Furthermore, we show that the aggregate patterns in attitude change over time are principally due to forces other than changes in life-course structure. While decreased rates of marriage explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased rates of education across all age groups. Furthermore, much of the shift in aggregate attitude stability remains unexplained, suggesting broader social structural features such as increased access to information with which to form opinions, political and cultural polarization, and more are the key drivers of attitude stability.

@@

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on what Lizardo [-@lizardo2017] calls "personal culture" -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level. In political socialization, "generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@gerber1998; @mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@ryder; @fosse2023a; @elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These aggregate models of cultural imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time. At a basic level, the "impressionable years" hypothesis requires that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@]. These tests tend to find higher rates of attitude change among the youngest adults -- people ages 18-25 -- and little difference between adults of other age ranges. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains more variance than period, the year of observation, for most items in the General Social Survey, suggesting an impressionable-years pattern.

In recent years, researchers have also developed new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022]. This work has produced remarkably similar patterns to previous work with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019]. However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015 -- which has not been tested.

@@

While the empirical pattern of early openness followed by relative stability appears quite robust, its cause is not clear. At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for the observed age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both these processes change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

## Biological and Cognitive Mechanisms

@@

It is also possible that there is something unique about human cognition during these periods that makes people ascribe more weight to the information acquired within them than to information acquired at other points. Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015]. Supporting this model is evidence that early formative periods have been observed not just in people, but across species. For example, in a study of social learning among birds, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.

@@

The thrust of the biological and cognitive arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages, regardless of their social environments. However, while biological mechanisms can shape how people respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change. For that, we must turn attention to the structuring of the social world.

@@

Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct, characterized by sustained social learning and constant changes of social contexts. Here we briefly consider some of the ways that the "impressionable years" differ from other life stages that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.

First, people in this age group tend to move across social contexts at a much higher rates than people in other age groups. In the 11-year window from ages 14 to 25, in which people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information. People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons. Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups [NEED CITES HERE].

Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions. For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic2021]. Personal social networks also differ in this time period. Young peoples' networks tend to be larger and include more friends than kin [@marsden1987]. It could be the case that these larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].

@@

While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on people's attitudes is contingent on their ability to incorporate it into their cultural repertoire. The social conditions of different life stages might make it easier or more difficult for people to act on these new sources of information in ways that facilitate opinion change.

Social networks not only shape peoples' exposure to new information, but they can also affect the degree to which people adopt new beliefs and attitudes. People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs, even when they are exposed to new ideas [@stark1980]. Consistent with this, mid-life is associated with denser, more homogenous networks that include larger numbers of family [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004]. If young people exist in less dense, more heterogeneous social networks, it might be easier for them to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks. Similarly, if information flows less freely through weak-tie networks, these structures might allow respondents to limit their exposure to or close off parts of their lives to people who might not be supportive of new attitudes.

Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power, which tend to occur more frequently during middle age. As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information or forcing them to develop stable opinions on issues where previously they felt capable of ambivalence [@eaton2009]. This potentially explains another phenomenon: that people appear to open up to changing their minds on some issues as they enter later life and retirement, when these social expectations of stability presumably diminish [@danigelis2007].

Existing research suggests that the structure of social life during adolesence, early adulthood, and later life stages shape both exposure to new information and a person's ability to update attitudes, opinions, and beliefs in the face of that information. This further suggests that changes in these life course stages should matter for the amount of opinion change we observe.

@@

@@

First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for successive cohorts as they pass through the same ages. Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce. In the United States, both the median age at first marriage and median age of first childbirth has consistently risen since the mid-20th Century [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts.

Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce or movement back into education after several years in the workforce. These transitions are also accompanied by heightened rates of geographic mobility, as well as movement between the workforce and education.

If the distinctive nature of "impressionable years" is principally driven by their distinct social circumstances, then these shifts have the potential to extend these social circumstances into later ages. This, in turn, could have an effect on how long exploration, and therefore openness to change, lasts throughout the life-course. If this is the case, we should expect to see heightened levels of attitude change among age groups that, for earlier cohorts, displayed mainly stability. In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age. Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to older adults in their rates attitude change but that subsequent cohorts will diverge from older adults, making more change over time.

*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 18 to 25 than cohorts born in earlier years.*

We expect a similar but more muted trajectory for the attitude behavior of people aged 25-33. Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research [@krosnic1991; @alwin1991], but that more recent cohorts will come to more closely resemble the 18-25 age group, diverging from middle aged respondents.

*Hypothesis 2: More recent cohorts will show higher rates of attitude change in the age range of 26 to 33 than cohorts born in earlier years.*

Consistent with life-course structuring theories, we expect these divergences will principally be explained by changes in the timing of life course transitions such as marriage and childbirth. Controlling for these life-course transitions should eliminate the gap between the youngest adults and later adults.

*Hypothesis 3: Controlling for rates of marriage and childbirth, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.*

@@

Rosenfeld's work suggests that the social conditions that shape where people spend their impressionable years might also shape how much cultural exploration they can engage in. An increase in independence might entail more exploration and greater ability to update attitudes as a result of this exploration. This assertion in turn engenders a second prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.

@@

*Hypothesis 4: Controlling for life course transitions, more recent cohorts of early adults (18-33) will make higher rates of attitude change than contemporaneous middle-aged adults.*

@@

Finally, while we have primarily focused on changes in the timing and structure of life course stages over the past 70 years, that time frame has also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure. We briefly consider three broader changes here.

First, the number of people completing high school and pursuing and completing higher levels of education has increased significantly over the window studied here [@schofer2005; @hout2012]. Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007]. The specific mechanism that links educational attainment and attitude stability is unclear, with some perspectives suggesting educational attainment improves cognitive function and memory while other perspectives suggest that educational attainment makes people pay more attention to current events. It could also be the case that education cultivates in people a normative expectation that they should hold opinions. Regardless of the specific mechanism, a population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.

Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding. Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes opinion-holding easier. To the extent that stable opinions are in part a reflection of signals from elites [@converse1965; @zaller1992], then we should expect these shifts to stabilize opinions across all age ranges.

Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008]. People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta [-@dellaposta2020] calls an "oil spill" of polarization. While clarity around political issue positions is most commonly studied, the same forces can also align other social beliefs and identities. This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992]. Geographic and social sorting into lifestyle "enclaves" also reinforce these processes, continually providing clearer signals about which opinions people should hold [@dellaposta2015].

Broadly speaking, these three forces suggest that opinions should broadly become more stable over time, regardless of life-course stages. If this is the case, we might not observe an absolute decrease in stability for early adults, but rather a relative decrease, as extension of the life course works to decrease stability for members of these age groups while broader social forces stabilize opinions for all groups.

@@

Testing our hypotheses requires data on opinion and attitude reports over time for members of different cohorts in the same age ranges, with respondents coming from the same general population. We also require information about life-course transitions standardized across these panels. To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years. Both the ANES and GSS are designed to generalize to the non-institutionalized U.S. adult (18+) population, and therefore can be thought of as tapping the same general population. We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20. We evaluated an additional panel conducted by the ANES in 2008-10, but no questions overlapped this panel and any other panel.
